---
title: "Midterm Advanced Modelling: A PCA and Clustering approach to Degrowth Theory"
format: html
editor: visual
---

## 0. Setup

```{r}
library(eurostat)
library(tidyverse)
library(mice)
library(knitr)
library(GGally)
library(factoextra)
library(sf)
library(fmsb)
library(cluster)

set.seed(123)
```

## 1. Introduction

### TO FIX Motivation and Research Question

“All living organisms grow. But in nature there is a self-limiting logic to growth (...) **When growth fails to stop – when cells keep replicating just for the sake of it – it's because of a coding error, like what happens with cancer**.” — Jason Hickel, Less is More

The concept of degworth (or eco-socialism) as en economic system has gained more and more importance mostly in academic circles but also in policymaking as a response to the growing notion that an ever-increasing economic growth is not feasible with a finite planet with finite resources. Degrowth differs from "sustainable growth" or "green growth" narratives because they part ways at the level of principles: they differ fundamentally in their understandings of capitalism, economic growth, and the political mechanics of societal change. The sustainable growth worldview rests on coherence: the belief that crises stem from policy mistakes, missing data, misaligned incentives. Fix the context and the system will behave. Degrowth rests on dialectics: crises arise from the system itself—its structure, its power, its hunger for expansion. One approach wants to coordinate interests. The other confronts domination. They do not just diverge on methods; they disagree on reality.

This has concrete implications that fundamentally change the material conditions of a society. In Europe, we have seen how certain policies such as the Green Deal, and the Just Transition Fund are tackling some how these changes. That said they are more focused on "green growth" rather than real Degrowth. In addition, not all European regions are created equal and are equally positioned to navigate towards this transition. Regions differ in social protection systems, diversified economic structures, and existing ecological embeddedness. Understanding this big variation is essential for policymakers that are interested in real Degrowth.

This project intends to answer this research question: **What structural profiles characterize European NUTS-2 regions in terms of their "readiness" for a post-growth transition, and can we identify meaningful typologies of degrowth "readiness" across the continent?**

In order to answer this research question, I will apply Principal Component Analysis (PCA) to reduce a multidimensional set of regional indicators into interpretable components capturing the main drivers of variation in degrowth readiness. I will then apply hierarchical and K-means clustering to group regions into structural typologies. The resulting clusters are then externally validated against life expectancy and carbon emissions to assess whether the components capture meaningful socioeconomic and ecological differences.

### 1.2 Theoretical Framework

"Degrowth readiness" is understood here as a construct with three main theory-motivated pillars.

The first pillar is social protection capacity: the extent to how can a region' institutions and economic structure can absorb a potential contraction in GDP without reversing it on top of its population. This pillar lays its foundations on the degrowth literature's emphasis on unconditional and uncompromising provision of basic services as preconditions for any viable post-growth scenario.

The second pillar is economic structure: the extent to which a region's productive base is oriented toward high-output, resource-intensive sectors versus more local and lower-intensity activities. Regions heavily dependent on manufacturing industry face greater structural adjustment costs in a degrowth transition, while regions with diversified service economies or significant agricultural sectors may have greater flexibility. That is why I added R&D expenditure as it could be an indicator that could be a driver in singlaing green innovation or reliance on intensive and extractive industries.

The third pillar is ecological embeddedness: the extent to which a region already operates within lower-footprint production and consumption patterns. The hypothesis is that regions that already score well on these parameters have materially kick-started the material transition that degrowth requires, regardless of whether this reflects deliberate policy or structural factors.

## 2. Data

### 2.1 Overview

The analysis uses Eurostat regional data at NUTS-2 level. This is the standard EU subnational unit. Each regional unit typically contains 800,000 to 3 million inhabitants. I have chosen to use 2021 as the reference year because:

1)  it was the most complete year available
2)  it also includes geographical european but non-EU countries like Norway or Turkey for example.

That said, there may be a little limitation to this year as 2021 falls right in the post-COVID recovery period. Therefore, some of the variables used here such as, employment composition, education, and ecological indicators may be affected by this. However, they are slow-moving enough as variables that this does not materially affect the analysis to a great extent. More cyclically sensitive variables like unemployment and GDP are kept with this limitation in mind.

### 2.2 Geographic Scope

As aforementioned, instead of restricting the analysis to EU-27, I adopt a broader European scope covering all countries for which Eurostat publishes NUTS-2 data.

```{r}
europe_scope <- c(
  # EU-27
  "AT","BE","BG","CY","CZ","DE","DK","EE","EL","ES",
  "FI","FR","HR","HU","IE","IT","LT","LU","LV","MT",
  "NL","PL","PT","RO","SE","SI","SK",
  # Non-EU
  "NO",  # Norway
  "CH",  # Switzerland
  "TR",  # Turkey
  "RS",  # Serbia
  "ME",  # Montenegro
  "MK",  # North Macedonia
  "AL",  # Albania
  "GE",  # Georgia
  "AM",  # Armenia
  "AZ",  # Azerbaijan
  "MD"   # Moldova
)
```

This wider geographic range provides more meaningful variation for clustering. All datasets are filtered to NUTS-2 level by retaining only geographic codes of exactly four characters, using the helper function:

```{r}
nuts2 <- function(df) df |> filter(nchar(geo) == 4)
```

### 2.3 Variables

I collect 15 variables across three degrowth pillars: Social protection (7 variables): GDP per capita in PPS (nama_10r_3gdp), unemployment rate (lfst_r_lfu3rt), life expectancy at birth (demo_r_mlifexp), household disposable income (tgs00006), poverty risk rate (ilc_peps11n), early school leaving (edat_lfse_16), and tertiary education attainment (edat_lfse_04).

```{r}
# 1. GDP PER CAPITA
gdp <- get_eurostat("nama_10r_3gdp", time_format = "num") |>
  filter(unit == "EUR_HAB", TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, gdp_pps = values)
cat("gdp:", nrow(gdp), "rows\n")

# 2. UNEMPLOYMENT RATE
unemp <- get_eurostat("lfst_r_lfu3rt", time_format = "num") |>
  filter(age == "Y15-74", sex == "T", isced11 == "TOTAL",
         unit == "PC", TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, unemployment = values)
cat("unemp:", nrow(unemp), "rows\n")

# 3. LIFE EXPECTANCY
life_exp <- get_eurostat("demo_r_mlifexp", time_format = "num") |>
  filter(sex == "T", age == "Y_LT1", TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, life_expectancy = values)
cat("life_exp:", nrow(life_exp), "rows\n")

# 4. DISPOSABLE INCOME
income <- get_eurostat("tgs00006", time_format = "num") |>
  filter(TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, disposable_income = values)
cat("income:", nrow(income), "rows\n")

# 5. AT-RISK-OF-POVERTY
poverty <- get_eurostat("ilc_peps11n", time_format = "num") |>
  filter(TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, poverty_risk = values)
cat("poverty:", nrow(poverty), "rows\n")

# 6. EARLY SCHOOL LEAVING
esl <- get_eurostat("edat_lfse_16", time_format = "num") |>
  filter(sex == "T", TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, early_school_leaving = values)
cat("esl:", nrow(esl), "rows\n")

# 7. TERTIARY EDUCATION
tertiary <- get_eurostat("edat_lfse_04", time_format = "num") |>
  filter(sex == "T", age == "Y25-64",
         isced11 == "ED5-8", TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, tertiary_edu = values)
cat("tertiary:", nrow(tertiary), "rows\n")
```

Economic structure (5 variables): Employment shares in agriculture, industry and services (lfst_r_lfe2en2), and R&D expenditure as % of GDP (rd_e_gerdreg). Population density (demo_r_d3dens) is also included here as a structural context variable separating capital city regions from peripheral ones.

```{r}
# ── 8. EMPLOYMENT IN AGRICULTURE (convert to %) ──────────────────────────────
emp_total <- get_eurostat("lfst_r_lfe2en2", time_format = "num") |>
  filter(sex == "T", age == "Y15-64",
         nace_r2 == "TOTAL", TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, emp_total = values)

emp_agri <- get_eurostat("lfst_r_lfe2en2", time_format = "num") |>
  filter(sex == "T", age == "Y15-64",
         nace_r2 == "A", TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, emp_agri_abs = values) |>
  left_join(emp_total, by = "geo") |>
  mutate(emp_agriculture = (emp_agri_abs / emp_total) * 100) |>
  select(geo, emp_agriculture)
cat("emp_agri:", nrow(emp_agri), "rows\n")

# ── 9. EMPLOYMENT IN INDUSTRY (convert to %) ─────────────────────────────────
emp_industry <- get_eurostat("lfst_r_lfe2en2", time_format = "num") |>
  filter(sex == "T", age == "Y15-64",
         nace_r2 == "B-E", TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, emp_ind_abs = values) |>
  left_join(emp_total, by = "geo") |>
  mutate(emp_industry = (emp_ind_abs / emp_total) * 100) |>
  select(geo, emp_industry)
cat("emp_industry:", nrow(emp_industry), "rows\n")

# ── 10. EMPLOYMENT IN SERVICES (convert to %) ────────────────────────────────
emp_services <- get_eurostat("lfst_r_lfe2en2", time_format = "num") |>
  filter(sex == "T", age == "Y15-64",
         nace_r2 == "G-I", TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, emp_serv_abs = values) |>
  left_join(emp_total, by = "geo") |>
  mutate(emp_services = (emp_serv_abs / emp_total) * 100) |>
  select(geo, emp_services)
cat("emp_services:", nrow(emp_services), "rows\n")

# 11. R&D EXPENDITURE
rd_exp <- get_eurostat("rd_e_gerdreg", time_format = "num") |>
  filter(unit == "PC_GDP", sectperf == "TOTAL",
         TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, rd_expenditure = values)
cat("rd_exp:", nrow(rd_exp), "rows\n")

# 12. POPULATION DENSITY
pop_density <- get_eurostat("demo_r_d3dens", time_format = "num") |>
  filter(TIME_PERIOD == 2021) |>
  nuts2() |>
  select(geo, pop_density = values)
cat("pop_density:", nrow(pop_density), "rows\n")
```

Ecological embeddedness (3 variables):Greenhouse gas emissions (env_ac_ainah_r2), municipal waste per capita (env_wasmun), and renewable energy share (nrg_ind_ren). These three variables are only available at national level in Eurostat and are therefore assigned uniformly to all regions within a country. This is an acknowledged limitation — the ecological pillar captures national policy rather than regional variation.

```{r}
# 13. GHG (country-level, joined to NUTS-2)
ghg <- get_eurostat("env_ac_ainah_r2", time_format = "num") |>
  filter(airpol == "GHG", nace_r2 == "TOTAL",
         unit == "T", TIME_PERIOD == 2021,
         nchar(geo) == 2) |>
  select(country = geo, ghg_emissions = values)
cat("ghg:", nrow(ghg), "rows\n")

# 14. WASTE (country-level, joined to NUTS-2)
waste <- get_eurostat("env_wasmun", time_format = "num") |>
  filter(unit == "KG_HAB", wst_oper == "GEN",
         TIME_PERIOD == 2021, nchar(geo) == 2) |>
  select(country = geo, municipal_waste = values)
cat("waste:", nrow(waste), "rows\n")

# 15. RENEWABLE ENERGY (country-level, joined to NUTS-2)
renewables <- get_eurostat("nrg_ind_ren", time_format = "num") |>
  filter(nrg_bal == "REN", unit == "PC", TIME_PERIOD == 2021) |>
  filter(nchar(geo) == 2) |>
  select(country = geo, renewables_share = values)
cat("renewables:", nrow(renewables), "rows\n")
```

### 2.4 Cleaning and Imputation

The initial joined dataset contained 285 NUTS-2 regions.

```{r}
final_data <- gdp |>
  left_join(unemp,        by = "geo") |>
  left_join(life_exp,     by = "geo") |>
  left_join(income,       by = "geo") |>
  left_join(poverty,      by = "geo") |>
  left_join(esl,          by = "geo") |>
  left_join(tertiary,     by = "geo") |>
  left_join(emp_agri,     by = "geo") |>
  left_join(emp_industry, by = "geo") |>
  left_join(emp_services, by = "geo") |>
  left_join(rd_exp,       by = "geo") |>
  left_join(pop_density,  by = "geo") |>
  mutate(country = substr(geo, 1, 2)) |>
  left_join(renewables,   by = "country") |>
  left_join(ghg,          by = "country") |>
  left_join(waste,        by = "country")
```

However, it was easy to spot that there were missingness issues.

```{r}
print(sapply(final_data, function(x) round(sum(is.na(x)) * 100 / nrow(final_data), 1)))
```

Here I can see the percentage of missingness per variable

```{r}
# Which regions are missing poverty_risk?
final_data |>
  filter(is.na(poverty_risk)) |>
  mutate(country = substr(geo, 1, 2)) |>
  count(country) |>
  arrange(desc(n))
```

```{r}
# Which regions are missing early_school_leaving?
final_data |>
  filter(is.na(early_school_leaving)) |>
  mutate(country = substr(geo, 1, 2)) |>
  count(country) |>
  arrange(desc(n))
```

```{r}
# Which regions are missing renewables_share?
final_data |>
  filter(is.na(renewables_share)) |>
  mutate(country = substr(geo, 1, 2)) |>
  count(country) |>
  arrange(desc(n))
```

```{r}
# Overall: how many regions are missing 3 or more variables?
final_data |>
  mutate(n_missing = rowSums(is.na(across(gdp_pps:municipal_waste)))) |>
  count(n_missing) |>
  arrange(n_missing)
```

```{r}
final_data |>
  mutate(n_missing = rowSums(is.na(across(gdp_pps:municipal_waste)))) |>
  filter(n_missing >= 5) |>
  select(geo, n_missing, everything()) |>
  arrange(desc(n_missing))
```

Therefore, three cleaning decisions were applied.

First, the three Albanian regions were dropped due to near-total missingness (9 out of 15 variables missing each). The Åland Islands and Mayotte were also removed because they are micro-territories with unrepresentative profiles. Second, Turkey was excluded entirely. It could seem as an abrupt decision as it is a big country with a lot of interesting data, but Turkish regions were missing both poverty risk and renewable energy share. In addition, Turkey's development trajectory substantially differs from the European context and therefore there was the risk of substantially skew the analysis.

```{r}
# Drop Turkey, Albania, and Åland Islands
final_data_clean <- final_data |>
  filter(
    !substr(geo, 1, 2) %in% c("TR", "AL"),  # Drop Turkey and Albania
    !geo %in% c("FI20", "FRY5")              # Drop Åland and Mayotte
  )
```

After exclusions the dataset contained 254 regions. Remaining missingness was concentrated in poverty risk (16%, mainly France which does not report this at NUTS-2), early school leaving (20%, spread across multiple countries), R&D expenditure (17%, smaller regions tend not to report this), and emp_agriculture (13% as urban areas tend not to report it). All other variables were below 6.5% missing, with most of them very close at 0%.

```{r}
print(sapply(final_data_clean, function(x)
  round(sum(is.na(x)) * 100 / nrow(final_data_clean), 1)))
```

Remaining missing values were habdled with imputation using a random forest algorithm through the mice package, with four imputation cycles and a fixed seed for reproducibility. This preserves the multivariate correlation structure of the data, following the same approach used in the course reference notebooks.

```{r}
# Prepare matrix for imputation (exclude identifier columns)
data_for_imputation <- final_data_clean |>
  select(-geo, -country)

# Run mice with random forest (same method as course template)
set.seed(123)
mice_mod <- mice(data_for_imputation, m = 4, method = "rf",
                 printFlag = FALSE)

# Complete the dataset
data_imputed <- complete(mice_mod, action = 4)

# Add identifiers back
data_imputed$geo     <- final_data_clean$geo
data_imputed$country <- final_data_clean$country

# Verify no missing values remain
print(sapply(data_imputed, function(x) sum(is.na(x))))
```

The final dataset contains 254 NUTS-2 regions with complete information on all 15 variables.

```{r}
head(data_imputed)
```

## 3. Pre-processing and Descriptive Analysis

### 3.1 Data Loading

I begin by loading the cleaned and imputed dataset and preparing it for analysis.

```{r}
# Load the clean dataset
data <- read.csv("degrowth_nuts2_2021.csv")
```

### 3.2 Log Transformations of GDP and disposable income

GDP per capita and disposable income per capita show a right-skewed distribution to this type of data, where a small number of wealthy regions and individuals pull the tail. Therefore, I apply here a log transformation to these variables, because otherwise they would highly influence the principal components.

As I can see from the transormations from the before and after I have a much more symmetric distribution:

```{r}
#GDP per capita
par(mfrow = c(1, 2))
hist(data$gdp_pps, breaks = 30, main = "GDP (Original)", 
     col = "lightcoral", xlab = "GDP per capita (EUR)")
hist(log(data$gdp_pps), breaks = 30, main = "GDP (Log-transformed)", 
     col = "lightgreen", xlab = "log(GDP)")
par(mfrow = c(1, 1))
```

```{r}
#Disposable Income
par(mfrow = c(1, 2))
hist(data$disposable_income, breaks = 30, main = "Disposable Income", 
     col = "lightcoral", xlab = "GDP per capita (EUR)")
hist(log(data$disposable_income), breaks = 30, main = "Disposable Income (Log-transformed)", 
     col = "lightgreen", xlab = "log(GDP)")
par(mfrow = c(1, 1))
```

Here, I prepare the data matrix for PCA and exclude the identifier columns `geo` and `country`

```{r}
# Prepare data matrix for PCA
X <- data |>
  select(-geo, -country) |> 
  mutate(gdp_pps = log(gdp_pps),
         disposable_income = log(disposable_income))
```

### 3.3 Summary Statistics

```{r}
# Generate summary statistics
summary_stats <- X |>
  summarise(across(everything(), 
                   list(Mean = ~round(mean(.), 2),
                        SD = ~round(sd(.), 2),
                        Min = ~round(min(.), 2),
                        Max = ~round(max(.), 2)))) |>
  pivot_longer(everything(), 
               names_to = c("Variable", ".value"), 
               names_sep = "_(?=Mean|SD|Min|Max)")

kable(summary_stats, 
      caption = "Summary Statistics for 15 Variables (n=254 regions)")
```

The first impression of this Summary Statistics is that there is a very high heterogeneity across European regions.

-   GDP per capita (logged) ranges from 8.58 to 11.64, corresponding to roughly €5,300 to €113,000 in original units which is a 20x difference between the poorest and wealthiest regions.

-   Unemployment varies from 1.8% to 30.8%, which could be a reflection of two main aspects: structural differences in labor market and the uneven post-COVID recovery of 2021.

-   Life expectancy spans 69.5 to 85.2 years, indicating profound disparities.

-   Poverty risk averages 22% but ranges from 6% to 50%, suggesting that even within wealthy European countries, significant pockets of deprivation persist.

-   Educational shows wide variation: early school leaving ranges from under 1% to over 23%. Tertiary education varies from 12% to 61%. This tells us that education policies are very different from region to region.

-   Employment structure reveals the expected rural-urban divide. Agriculture accounts for just 5% of employment on average but reaches 26% in peripheral rural regions, while services average 23% but range up to 52% in metropolitan areas dominated by tertiary sector activity. Industry employment is more evenly distributed (mean 18%, range 4-39%), though with notable concentrations in traditional manufacturing belts.

In addition, certain variables such as GHG emissions, municipal waste and population density showcase nominal numbers ad therefore highlight the need of standardization before PCA, ensuring variables contribute based on correlation patterns rather than arbitrary measurement units.

### 3.4 Variable Distributions Before and After Standardization

The following graphs tells us why standardization is absolutely necessary for this analysis:

```{r}
# Original scale
boxplot(X, las = 2, col = "lightblue", 
        main = "Original Scale",
        cex.axis = 0.6, cex.main = 0.9)
```

```{r}
# Standardized scale
X_scaled <- scale(X)
boxplot(X_scaled, las = 2, col = "steelblue", 
        main = "Standardized (mean=0, sd=1)",
        cex.axis = 0.6, cex.main = 0.9)
```

GHG emissions was dominating the chart while after standardization all variables are centered at zero with unit variance, ensuring that PCA treats each variable equally and extracts components based on correlation patterns rather than measurement units.

### 3.5 Correlation Structure

```{r}
ggcorr(X, label = TRUE, label_size = 1.5, hjust = 0.92, 
       layout.exp = 2) +
  labs(title = "Correlation Matrix of Degrowth Variables")
```

The correlation matrix reveals clear clustering patterns that justify the use of PCA. A strong **"social development"** bloc emerges: GDP correlates strongly with disposable income (r = 0.9) and life expectancy (r = 0.7), while showing a negative correlations with poverty risk (r = -0.4). Tertiary education fits into this bloc with positive correlations to GDP (r = 0.6), disposable income (r = 0.6), and life expectancy (r = 0.4). Unemployment displays a distinctive pattern: it correlates positively with poverty risk (r = 0.6) but shows essentially no correlation with GDP (r = 0.1), suggesting that labour markets operate differently region by region and that are somewhat independent of the overall regional wealth. This coherent pattern suggests these variables tap into a common underlying dimension of regional prosperity and human capital.

Regarding the **employment structure** they show the expected trade-offs. Agriculture correlates negatively with services (r = -0.4), tertiary education (r = -0.4), GDP (r = -0.3), and positively with poverty risk(r = 0.5) and unemployment (r = 0.4); telling us that this is a seasonal industry and with people often working only part of the year and without many opportunities for wealth creation. Industry employment shows moderate negative correlation with services (r = -0.4) and tertiary education (r = -0.4), suggesting traditional manufacturing regions have distinct profiles from knowledge economy hubs. In addition it also negatively correlates (r = -0.4) with life_expectancy, meaning that hard manual work is likely a driving factor in lowering years of life expectancy. R&D expenditure correlates most strongly with tertiary education (r = 0.5) and municipal waste (r = 0.5), potentially flagging wealthy, innovation-intensive regions with high consumption patterns.

Out of the three **ecological variables** — renewables, GHG emissions, and municipal waste — only the latter one shows correlations with socioeconomic indicators. This could potentially be due to their country-level aggregation (which washes out within-country regional variation) and genuine independence of environmental performance from standard development metrics. Notably, renewables share shows weak negative correlation with disposable income (r = -0.1) and population density (r = -0.1), suggesting that renewable energy adoption is not systematically higher in wealthier or more urbanized regions.

In general, the matrix shows a **medium correlation structure** (many r \> 0.5) and without problematic multicollinearity issues only gdp and disposable income with a (r \> 0.9), making the data well-suited for PCA. The clear emergence of a correlation block of social development, employment structure, and weakly-related ecological indicators hint at the most likely principal component I expect.

## 4 Principal Component Analysis

In this section I will carry out a Principal Component Analysis (PCA) which will reduce the 15 variables into uncorrelated components, ordered by variance explained. At times, degrowth analysis faces a paradox: one of the harshest critiques degrowth scholars is that GDP is a fallacious matrix and does not reflect the true wellbeing of a nation. However, many analysis and this one will also be dominated by economic development indicators such as GDP per capita. This reflects degrowth theory's core insight—only societies achieving material sufficiency can afford to prioritize wellbeing over GDP. Regions fighting poverty and unemployment need growth; they cannot realistically "degrow."

However, degrowth involves more than material capacity. It also concerns ecological performance and economic structure. I therefore interpret three principal components capturing distinct aspects of degrowth readiness: material capacity (PC1), economic structure (PC2), and environmental performance (PC3).

### 4.1 PCA Execution

```{r}
# Run PCA on standardized data
pca <- prcomp(X, scale. = TRUE)

# Basic summary
summary(pca)
```

The first principal component captures 34% of total variance, the second 15%, and the third 11%, bringing cumulative explained variance to **60% with three dimensions**. PC4 adds 8%, such that the first four components explain 68%.

### 4.2 Variance Explained

```{r}
fviz_screeplot(pca, addlabels = TRUE, ncp = 10) +
  labs(title = "Scree Plot: Variance Explained by Principal Components",
       x = "Principal Component",
       y = "Percentage of Variance Explained")
```

In this plot I visually see what was explained in the previous section

### 4.3 Component Loadings

```{r}
# Extract and display loadings for first three components
loadings <- pca$rotation[, 1:3]
kable(round(loadings, 3), 
      caption = "Variable Loadings on First Three Principal Components")
```

**PC1** shows clear bipolar structure: wealth indicators (GDP, income, tertiary education, R&D) cluster with negative signs, while economic vulnerability (agriculture, poverty, unemployment) clusters with positive signs. This is synonym of an economic development gradient.

**PC2** is dominated by industrial employment (+0.55), with unemployment, service employment, and poverty being negative and therefore capturing economic/industrial structure independent of development level.

**PC3** is defined by GHG emissions (-0.59) and early school leaving (-0.52) being negative, and with renewables share (+0.31) being positive; signalling an environmental-educational dimension.

**PC4** (8%) appears driven by renewables_share and pop_density but is not interpreted in detail. Together, the first four components explain 68% of variance, but I will focus on PC1-PC3 for interpretability.

### 4.4 Interpretation of Principal Components

#### 4.4.1 PC1: Socioeconomic Development / Material Capacity for Degrowth (34%)

The first component is dominated by development indicators. At the strongest negative values we have: GDP per capita (-0.42), disposable income (-0.40), tertiary education (-0.31), R&D expenditure (-0.32), life expectancy (-0.28), municipal waste (-0.29). On the positive end: agricultural employment (+0.32), poverty risk (+0.27), unemployment (+0.20).

```{r}
fviz_contrib(pca, choice = "var", axes = 1, top = 15) +
  labs(title = "Variable Contributions to PC1",
       subtitle = "Socioeconomic Development Dimension")
```

**Degrowth Interpretation:** Regions with negative PC1 scores (which I invert for intuitive visualization) have solved fundamental material problems such as low poverty, low unemployment, universal education and healthcare. These regions possess degrowth capacity to reduce GDP while maintaining a high quality of life if policies were channeled through an equal re-distribution.

On the other hand, regions with positive PC1 scores face disadvantages making degrowth materially almost impossible at the present moment. A region with 25% poverty, 15% unemployment, and 40% agricultural employment cannot "degrow" without running into catastrophic consequences. These regions need conventional development, diversification, and educational expansion.

**The Paradox:** Regions scoring highest have the largest ecological footprints. Wealthy Northern European regions consume vastly more resources, generate more waste, contribute more to climate change than poor Eastern European regions. They are champions of overconsumption, not current degrowth practitioners. Yet this is exactly what degrowth theory predicts. Degrowth advocates argue rich regions should consume less while maintaining wellbeing. PC1 identifies which regions could realistically do this by measuring distance from subsistence, thickness of buffers, availability of alternatives to growth-dependent welfare.

For visualization, I multiply PC1 by -1 so high values = high capacity.

#### 4.4.2 PC2: Economic Structure / Industrial Transition Challenges (15%)

The second component distinguishes regions by economic structure, closely tightened to development level. Industrial employment (+0.55) dominates the positive pole. Negative loadings: unemployment (-0.48), service employment (-0.31), poverty risk (-0.29), life expectancy (-0.28).

```{r}
fviz_contrib(pca, choice = "var", axes = 2, top = 15) +
  labs(title = "Variable Contributions to PC2",
       subtitle = "Economic Structure Dimension")
```

High PC2 regions are manufacturing heartlands: automotive clusters in Southern Germany, industrial zones in Northern Italy, Czech Republic, Poland. PC2's link to PC1 means these can be wealthy (Bavaria) or developing (Eastern Europe, southern Italy) at the same time without being mutually exclusive.

Negative PC2 captures two different types: prosperous service-based urban centers (London, Stockholm) that transitioned to post-industrial economies, and deindustrialized regions where factories closed but services never replaced lost jobs. Both are "non-industrial" but for opposite reasons.

**Degrowth Implications:** Heavy manufacturing regions face disruptive shifts requiring energy-intensive production, making it hard for them to be transitioning to circular models and potentially accepting reduced output without exterior input. However, they possess industrial capacity and engineering expertise for carrying out this change. Service economies may transition more easily (less materially bound) but face vulnerabilities if degrowth policies cuts spending on tourism, entertainment, luxury services, or speculative finance without properly redirecting jobs that will be displaced.

#### 4.4.3 PC3: Environmental Performance / Sustainability Orientation (11%)

The third component reveals environmental-educational quality independent of development and structure. Dominant loadings: GHG emissions (-0.59), early school leaving (-0.52), poverty risk (-0.31), renewables share (+0.31).

```{r}
fviz_contrib(pca, choice = "var", axes = 3, top = 15) +
  labs(title = "Variable Contributions to PC3",
       subtitle = "Environmental Performance Dimension")
```

High PC3 regions successfully adopted renewable energy and maintain better educational retention: Scandinavian regions with hydropower/wind, rural areas with renewable potential, regions managing successful green transitions. Low PC3 regions combine high emissions with educational problems: polluted industrial regions, car-dependent sprawl, fossil fuel-dependent areas.

**Why This Matters:** PC3 captures **current sustainability orientation**: the regions that already perform better on ecological metrics regardless of development level. This addresses PC1's limitation: a wealthy region with massive emissions (high PC1, low PC3) has material capacity but lacks ecological groundwork for smooth transition. Degrowth requires both wealth to afford GDP reduction AND movement toward sustainable practices.

### 4.5 Biplot: PC1 vs PC2

```{r}
fviz_pca_biplot(pca, 
                label = "var",
                col.var = "contrib",
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                repel = TRUE,
                pointsize = 1,
                alpha.ind = 0.3) +
  labs(title = "PCA Biplot: PC1 (Development) vs PC2 (Industrial Structure)")
```

The biplot displays regions (points) and variables (arrows) in PC1-PC2 space. Wealth variables (GDP, income, tertiary education, R&D, life expectancy) point left, forming a tight bundle defining development. Industrial employment points up, poverty and unemployment point right (low development). Environmental variables (GHG, renewables) have less of an effect here because they load more strongly on PC3.

### 4.6 Additional Biplots

```{r}
# PC1 vs PC3: Development vs Environmental Performance
fviz_pca_biplot(pca, 
                axes = c(1, 3),
                label = "var",
                col.var = "contrib",
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                repel = TRUE,
                pointsize = 1,
                alpha.ind = 0.3) +
  labs(title = "PCA Biplot: PC1 (Development) vs PC3 (Environmental Performance)")
```

This reveals how environmental performance crosscuts development. GHG emissions points down (low PC3 = high emissions), renewables share points up (high PC3 = high renewables), early school leaving points down. Crucially, environmental arrows are nearly perpendicular to wealth arrows—confirming independence. Lower-left panel being high development and poor environmental performance, while upper-right (low development, better environmental performance).

```{r}
# PC2 vs PC3: Industrial Structure vs Environmental Performance  
fviz_pca_biplot(pca, 
                axes = c(2, 3),
                label = "var",
                col.var = "contrib",
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                repel = TRUE,
                pointsize = 1,
                alpha.ind = 0.3) +
  labs(title = "PCA Biplot: PC2 (Industrial Structure) vs PC3 (Environmental Performance)")
```

Industrial employment (right) exactly allinged with the positive x-axis and and GHG emissions (down) alligned with the negative y-axis form an angle signalling industrial regions often have higher emissions, but not always (for example points that are in the upper-right paneñ). Renewables share points up-left—renewable adoption more common in non-industrial regions. Region scatter reveals diverse combinations.

### 4.7 Regional Rankings

```{r}
# Create PC scores (apply sign correction to PC1)
pc_scores <- data.frame(
  geo = data$geo,
  country = data$country,
  PC1 = -pca$x[, 1],  # Flip for intuitive interpretation
  PC2 = pca$x[, 2],
  PC3 = pca$x[, 3]
)

# Top 10 regions
cat("Top 10 Regions — Highest Socioeconomic Development (PC1):\n")
top_regions <- pc_scores |> 
  arrange(desc(PC1)) |>
  head(10) |>
  select(geo, country, PC1, PC2, PC3)
kable(top_regions, digits = 2)

# Bottom 10 regions
cat("\nBottom 10 Regions — Lowest Socioeconomic Development (PC1):\n")
bottom_regions <- pc_scores |> 
  arrange(PC1) |> 
  head(10) |>
  select(geo, country, PC1, PC2, PC3)
kable(bottom_regions, digits = 2)
```

Top-ranked regions by PC1 concentrate in Northern/Western Europe. These combine high incomes, educated populations, diversified economies, strong welfare systems. Looking across components: most top PC1 regions have moderate-to-negative PC2 meaning that have moved away from an industrial economy to one that is more service based. However, this is not the case for the two German regions who actually score the lowest in PC3, showing us how Germany is its own case in Europe as it is the industrial powerhouse of the continent, but we will see more on this later.

Lowest-ranked regions by PC1 in Eastern/Southern peripheries: Bulgarian, Romanian, Serbia, and North Macedonia. These share high poverty (\>25%), elevated unemployment (15-20%), agricultural dependence, low education, out-migration, weak institutions. Their PC2 scores are moderately high, but on average half of tbe German ones. Their PC3 scores are mixed—some have low emissions from low economic activity, some have higher scores especially RS21 in Serbia (PC2=0.62, PC3=1.48)

The gap spans \~9 standard deviations on PC1 which tells us that regions are ery different in Europe and they will require different kind of policies.

### 4.8 Geographic Distribution

```{r}
# Get NUTS-2 shapefile
nuts2_map <- get_eurostat_geospatial(
  resolution = "60",
  nuts_level = 2,
  year = 2021
)

# Join PC scores
map_data <- nuts2_map |>
  left_join(pc_scores |> select(geo, PC1, PC2, PC3), 
            by = c("NUTS_ID" = "geo"))
```

```{r}
# PC1 Map
map_pc1 <- ggplot(map_data) +
  geom_sf(aes(fill = PC1), color = "white", size = 0.1) +
  scale_fill_gradient2(
    low = "#d73027", mid = "#ffffbf", high = "#1a9850",     
    midpoint = 0, name = "PC1:\nDevelopment", na.value = "grey90"
  ) +
  coord_sf(xlim = c(-10, 35), ylim = c(35, 71)) +
  labs(title = "PC1: Socioeconomic Development") +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank(), plot.title = element_text(size = 10, hjust = 0.5))
map_pc1
```

**PC1 (Development):** Clear core-periphery gradient. Darker Green (high development) concentrates in Scandinavia, Benelux, Southern Germany, Austria, Ireland, Northern Italy and some areas of France. Isolated green islands in Eastern capitals (Prague, Warsaw, Budapest, Madrid, etc.) surrounded by red rural peripheries. Southern Europe mixed—coastal/metropolitan moderate-to-high, interior/islands low.

```{r}
# PC2 Map
map_pc2 <- ggplot(map_data) +
  geom_sf(aes(fill = PC2), color = "white", size = 0.1) +
  scale_fill_gradient2(
    low = "#2c7bb6", mid = "#ffffbf", high = "#d7191c",     
    midpoint = 0, name = "PC2:\nIndustrial", na.value = "grey90"
  ) +
  coord_sf(xlim = c(-10, 35), ylim = c(35, 71)) +
  labs(title = "PC2: Economic Structure") +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank(), plot.title = element_text(size = 10, hjust = 0.5))
map_pc2
```

**PC2 (Industrial Structure):** Red (industrial) regions form discontinuous band through Southern Germany, Northern Italy, Czech Republic, Poland—the manufacturing belt. A lot of Europe is of the neutral yellow. Blue (non-industrial) includes Scandinavian capitals, Western urban centers and souther regions of Italy, Spain and the whole Greece.

```{r}
# PC3 Map
map_pc3 <- ggplot(map_data) +
  geom_sf(aes(fill = PC3), color = "white", size = 0.1) +
  scale_fill_gradient2(
    low = "#8c510a", mid = "#f5f5f5", high = "#01665e",     
    midpoint = 0, name = "PC3:\nEnvironmental", na.value = "grey90"
  ) +
  coord_sf(xlim = c(-10, 35), ylim = c(35, 71)) +
  labs(title = "PC3: Environmental Performance") +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank(), plot.title = element_text(size = 10, hjust = 0.5))
map_pc3
```

**PC3 (Environmental Performance):**Teal (good performance) in all of teh Scandinavian regions (hydropower/wind), and in the Baltic countries plus some rural areas. Brown (poor performance) in some wealthy industrial areas (high emissions), car-dependent regions. Pattern doesn't mirror PC1 (wealth) or PC2 (industry)—confirms independence as here we do have Germany but we also have Southern Italy and Spain being in the lower end of PC1 and PC2 but scoring at the level of Germany here.

**Regional Types Emerging:** (1) Green-blue-teal regions (wealthy, service, good environment): Scandinavian urban areas—highest overall readiness. (2) Green-brown regions (wealthy, industrial, poor environment): German manufacturing—wealthy but environmentally challenged. (3) Red-mixed-brown regions (poor, mixed structure, poor environment): Some Eastern European regions—multiple disadvantages. (4) Red-mixed-teal regions (poor, low activity, decent environment): Rural areas with low emissions by default, not design.

Enormous within-country variation: Spain (Madrid/Barcelona green vs. rural red), Italy (Northern industrial vs. Southern rural), Romania/Bulgaria (capital islands in poverty seas), and so on.

### 4.9 Summary

**Key Findings:** PCA reduced 15 variables to three interpretable dimensions capturing 60% of variance. **PC1 (34%): Socioeconomic Development** represents material capacity for degrowth via post-scarcity logic—only regions solving basic material problems possess slack for GDP reduction. **PC2 (15%): Economic Structure** captures industrial intensity independent of development—revealing differential vulnerabilities in green transitions. **PC3 (11%): Environmental Performance** captures current ecological outcomes independent of wealth and structure—identifying which regions already perform better on sustainability metrics.

**Geographic Patterns:** Profound spatial heterogeneity. PC1 shows core-periphery (Northern/Western capitals vs. Eastern/Southern peripheries). PC2 identifies manufacturing belts. PC3 shows complex patterns uncorrelated with wealth—confirming policy choices matter independently. Gap between highest/lowest regions in the PC1 context spans 9 standard deviations—fundamentally different realities.

**Methodological Reflections:** 60% explained variance is respectable but indicates genuine multidimensionality—European regions differ beyond development/structure/environment and that perhaps the data we have available is not as thorough as it could be. The 40% unexplained reflects additional dimensions (urbanization, demographics, migration, specific industries, policy regimes, historical legacies). The three components represent **dominant systematic patterns** but don't completely satisfy the variation.

**Limitations:** (1) Capacity ≠ commitment—measuring material possibility, not political will. (2) PC1 essentially replicates HDI. I've labeled it "degrowth capacity" via theoretical argument, but empirical content is simply development. (3) Cross-sectional—misses dynamics and trajectories. (4) Environmental measures limited—only emissions/renewables, missing biodiversity, water, pollution. (5) 60% variance leaves substantial simplification.

**Transition to Clustering:** PCA identified continuous gradients, but not whether these structure discrete regional types or smooth continua. Visual inspection suggests natural groupings, but rigorous typology requires formal clustering analysis. Section 5 tests whether Europe's regional landscape is better characterized by continuous spectrum or discrete categories, incorporating information from higher components to capture more of the 40% unexplained variance.

## 5. Clustering Analysis

### 5.1 Introduction and Methodology

PCA revealed three dominant dimensions but left unresolved whether regions cluster into discrete types or distribute smoothly. This matters for policy: if degrowth readiness exists on a continuum, policies might scale smoothly; if discrete types exist, differentiated frameworks become necessary.

I employ hierarchical clustering to explore structure, then determine optimal K using three methods (elbow, silhouette, gap statistic), apply K-means for assignment, and validate using ANOVA.

```{r}
# Prepare data: use PC1-PC3 scores (60% variance, already standardized)
cluster_data <- pc_scores |> 
  select(PC1, PC2, PC3)
```

### 5.2 Hierarchical Clustering

```{r}
# Compute distance matrix and hierarchical clustering
dist_matrix <- dist(cluster_data, method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")

# Plot dendrogram with K=5 rectangles
plot(hc, labels = FALSE, hang = -1,
     main = "Hierarchical Clustering Dendrogram (Ward's Method)",
     xlab = "European NUTS-2 Regions", ylab = "Height (Dissimilarity)", sub = "")
rect.hclust(hc, k = 5, border = 2:6)
```

The dendrogram reveals a dominant bifurcation at height \~40, with two major super-clusters each subdividing further at heights \~23 and \~25, suggesting 5 natural groupings as shown by the colored boxes — consistent with the elbow and Gap statistic evidence. If the PCA loadings are as expected, the primary split likely reflects the wealthy/poor regional gradient (PC1), though this interpretation requires validation against cluster profiles.

### 5.3 Determining Optimal K

#### 5.3.1 Elbow Method

```{r}
# Compute WCSS for K = 1 to 10
set.seed(123)
wcss <- numeric(10)
for (k in 1:10) {
  km_temp <- kmeans(cluster_data, centers = k, nstart = 25)
  wcss[k] <- km_temp$tot.withinss
}

# Plot elbow curve
plot(1:10, wcss, type = "b", pch = 19, col = "steelblue",
     xlab = "Number of Clusters (K)", ylab = "Within-Cluster Sum of Squares",
     main = "Elbow Method for Optimal K")
grid()
```

The elbow plot shows no sharp inflection, but the curve's rate of decline visibly slows around K=5 or K=6, where WCSS reaches roughly 500–625. Either value is defensible as optimal; the choice should be guided by domain knowledge or supplementary metrics like the Silhouette score.

#### 5.3.2 Silhouette Analysis

```{r}
# Compute average silhouette width for K = 2 to 10
avg_sil <- numeric(9)
for (k in 2:10) {
  km_temp <- kmeans(cluster_data, centers = k, nstart = 25)
  sil <- silhouette(km_temp$cluster, dist_matrix)
  avg_sil[k-1] <- mean(sil[, 3])
}

# Plot silhouette scores
plot(2:10, avg_sil, type = "b", pch = 19, col = "darkgreen",
     xlab = "Number of Clusters (K)", ylab = "Average Silhouette Width",
     main = "Silhouette Method for Optimal K")
grid()
```

The Silhouette score peaks at K=3 (\~0.39), indicating this is where clusters are most cohesive and well-separated. The absolute value suggests weak-to-moderate structure which is quite typical for socioeconomic data with naturally fuzzy boundaries. K=3 is the clear optimum by this metric, conflicting with the elbow's suggestion of K=5/6 and asking us to make a judgment call based on interpretation or other metrics. Let's see if Gap statistics will help us.

#### 5.3.3 Gap Statistic

```{r}
# Compute gap statistic (comparison to null distribution)
set.seed(123)
gap_stat <- clusGap(cluster_data, FUN = kmeans, nstart = 25,
                    K.max = 10, B = 50)

# Plot and print optimal K
plot(gap_stat, main = "Gap Statistic Method for Optimal K")
print(gap_stat, method = "firstmax")
```

The Gap statistic rises steeply up to K=5/6 and plateaus thereafter, with overlapping error bars indicating no significant gain beyond K=6. Applying the standard one-SE rule, K=6 is the most defensible optimum as it has the smallest K whose Gap value falls within one standard error of the plateau.

#### 5.3.4 Synthesis

The three methods do not fully converge: the Silhouette score favors K=3, the Gap statistic suggests K=6, and the elbow is ambiguous between K=5/6 and the dendrogram supports 5 natural groupings. Noting that K=3 may undersplit meaningful regional variation while K=6 adds complexity without clear gain, I proceed with K=5 as a pragmatic compromise supported by the hierarchical structure.

### 5.4 K-Means Clustering

```{r}
# Run K-means with K=5
set.seed(123)
optimal_k <- 5

km_final <- kmeans(cluster_data, centers = optimal_k, 
                   nstart = 25, iter.max = 100)

# Add cluster assignments
pc_scores <- pc_scores |> mutate(cluster = km_final$cluster)

# Display cluster sizes
cat("Cluster Sizes:\n")
print(table(km_final$cluster))
cat("\nBetween-cluster variance / Total variance: ", 
    round(km_final$betweenss / km_final$totss * 100, 1), "%\n", sep="")
```

K-means produced five clusters of varying sizes (35–64 regions each), indicating reasonably balanced groupings. Between-cluster variance explains 72.3% of total variance, suggesting the five clusters capture meaningful structure in the data.

### 5.5 Cluster Visualization

```{r}
# 2D scatter: PC1 vs PC2
library(ggplot2)

ggplot(pc_scores, aes(x = PC1, y = PC2, color = factor(cluster))) +
  geom_point(size = 2.5, alpha = 0.7) +
  stat_ellipse(level = 0.95, size = 1.2) +
  scale_color_brewer(palette = "Set1", name = "Cluster") +
  labs(title = "Five-Cluster Solution in PC1-PC2 Space",
       x = "PC1: Development (34%)", y = "PC2: Industrial (15%)") +
  theme_minimal()
```

PC1 (development) drives primary separation: Cluster 2 occupies the high-industrial but still low-development left, Cluster 3 the low-industrial and low-development left, while Clusters 1, 4, and 5 concentrate on the right (higher development), differentiated vertically by PC2 (industrial structure). The overlapping ellipses for clusters 1, 4, and 5 aren't fuzzy boundaries due to noise; they reflect real gradations within Western European development models.

### 5.6 Cluster Characterization

```{r}
# PC profiles by cluster
cluster_pc_profiles <- pc_scores |>
  group_by(cluster) |>
  summarise(N = n(), PC1_Development = mean(PC1),
            PC2_Industrial = mean(PC2), PC3_Environmental = mean(PC3),
            .groups = "drop") |>
  arrange(desc(PC1_Development))

kable(cluster_pc_profiles, digits = 2,
      caption = "Mean Principal Component Scores by Cluster")
```

```{r}
# Original variable profiles
cluster_profiles_raw <- data |>
  mutate(cluster = pc_scores$cluster) |>
  group_by(cluster) |>
  summarise(N = n(), GDP = mean(gdp_pps, na.rm = TRUE),
            Poverty_Risk = mean(poverty_risk, na.rm = TRUE),
            Unemployment = mean(unemployment, na.rm = TRUE),
            Tertiary_Edu = mean(tertiary_edu, na.rm = TRUE),
            Emp_Industry = mean(emp_industry, na.rm = TRUE),
            GHG_Emissions = mean(ghg_emissions, na.rm = TRUE),
            Renewables = mean(renewables_share, na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(GDP))

kable(cluster_profiles_raw, digits = 1,
      caption = "Cluster Profiles: Key Variable Means")
```

------------------------------------------------------------------------

**Cluster 4 (N=64) -- Northern/Western core** (Purple on visualizations)

**PC profile:** PC1 = +2.35 (highest development), PC2 = -0.72 (mostly service-oriented), PC3 = +0.94 (good environmental performance)

**Key statistics:** GDP €49,118 (highest), poverty risk 16.8% (the lowest), unemployment 5.8%, tertiary education 43.9% (highest), industrial employment 13.2% (lowest), GHG emissions the lowest while having renewables the highest (34%)

The high PC1 combined with negative PC2 indicates post-industrial service economies. The strong PC3 score (highest among all clusters) reflects good environmental performance.

**Degrowth readiness:** Highest material wealth, robust safety nets, highly educated workforces, diversified service economies not tightly coupled to GDP growth. The strong environmental performance suggests many regions are already pursuing sustainability transitions. These are the regions that have both the economic strength and institutional frameworks to lead degrowth implementation and a transition towards an economy that serves the people and not growth just for the sake of it.

------------------------------------------------------------------------

**Cluster 1 (N=40) -- German industrial core** (Red on visualizations)

**PC profile:** PC1 = +1.62 (high development), PC2 = +1.35 (highly industrial), PC3 = -1.75 (poor environmental performance)

**Key statistics:** GDP €43,035 (second highest), poverty risk 20.6%, unemployment 3.6% (lowest), tertiary education 30.8%, industrial employment 22.4% (second highest), GHG emissions are at the highest and renewables lowest paired with cluster 5.

The combination of high PC1 (prosperity) with high PC2 (industrial intensity) and very low PC3 (poor environmental performance) reflects specialization in carbon-intensive manufacturing: automobiles, machinery, chemicals, steel.

**Degrowth readiness:** High material capacity but severe structural and environmental challenges. Prosperity is tightly linked to industrial production and with manufacturing employment (22.5%) nearly double the rate in Cluster 4. Highest emissions of all clusters indicate carbon intensity of production. However, these are regions that possess engineering expertise, skilled industrial workforce, and manufacturing infrastructure that could theoretically be repurposed for different economy models. The challenge is whether such transformation can preserve employment and regional identity built around industrial excellence.

------------------------------------------------------------------------

**Cluster 5 (N=62) -- The European Middle class** (Orange on visualizations)

**PC profile:** PC1 = +0.18 (slightly above average development), PC2 = -0.06 (mixed structure), PC3 = -0.02 (average environmental performance)

**Key statistics:** GDP €29,112 (middle), poverty risk 21%, unemployment 7.5%, tertiary education 33.3%, industrial employment 17.1%, GHG average and renewables the lowest paired with cluster 1.

The near-zero scores on all three PCs indicate these regions are genuinely average across development, industrial structure and environmental performance they define the European center of gravity and perhaps the true state of the EU at the moment.

**Degrowth readiness:** Ambiguous middle position. These regions have escaped extreme poverty and built basic infrastructure, placing them beyond Clusters 2-3, but haven't reached the post-scarcity conditions of Cluster 4. Here is where economic and industrial policy need to implement models that would have to work towards a degrowth focused economy where the wellbeing of the population and redistribution of capital and resources is the main focus and not something to be achieved by the classic trickle-down economic models. The overlapping confidence ellipse with Cluster 4 in PC1-PC2 space reflects that some Cluster 5 regions are converging toward core prosperity, while others still have significant ground to cover.

------------------------------------------------------------------------

**Cluster 3 (N=35) -- The Struggling Mediterrean** (Green on visualizations)

**PC profile:** PC1 = -2.41 (low development), PC2 = -2.26 (very much non-industrial), PC3 = -0.71 (below-average environmental performance)

**Key statistics:** GDP €18,340, poverty risk 33.1% (highest), unemployment 16.0% (highest), tertiary education 26.1% (second lowest), industrial employment 10.1 (lowest)%, GHG emissions average but surprisingly second highest renewables (20.2%)

The very negative PC2 (most non-industrial) combined with high unemployment distinguishes this from Cluster 2—these regions experienced deindustrialization or never industrialized, with service sectors (often tourism-dependent) failing to provide stable employment. Highest unemployment (16%) and high poverty (33.1%) indicate acute economic distress.

**Degrowth readiness:** Low capacity, compounded by structural challenges. These regions have already experienced involuntary "degrowth" through deindustrialization, fiscal austerity (especially Greece post-2010), and tourism sector volatility, without the wellbeing improvements degrowth theory promises. The combination of low GDP, high unemployment, and high poverty indicates severe material deprivation. Youth out-migration to Northern Europe further erodes human capital and demographic sustainability. In these regions, substantial investments need to be carried out to bring them to speed with the rest of Europe. However, the propensity that this region has towards renewables being the second highest cluster, tells us that there is possible adaptation for this and one of the potential industries where to focus.

------------------------------------------------------------------------

**Cluster 2 (N=53) -- The Eastern periphery** (Blue on visualizations)

**PC profile:** PC1 = -2.68 (lowest development), PC2 = +1.41 (industrial), PC3 = +0.68 (good environmental metrics)

**Key statistics:** GDP €12,509 (lowest by far), poverty risk 23.7% (second highest), unemployment 5.8% (second lowest), tertiary education 24.6% (lowest), industrial employment 26% (highest), GHG emissions and renewables are low.

The positive PC2 (unlike Cluster 3's negative) reflects remaining agricultural and light industrial employment rather than services. Paradoxically shows best PC3 (environmental metrics) alongside Cluster 4.

**Degrowth readiness:** Lowest material capacity. Degrowth is both materially and politically impossible in contexts of existing poverty (23.7%) and underdevelopment. These regions need conventional development: infrastructure investment (broadband, transport, healthcare facilities), educational expansion to retain youth, agricultural modernization, industrial diversification. The challenge is achieving this development sustainably—avoiding carbon-intensive industrialization pathways—while acknowledging that emissions will necessarily rise from current artificially low levels as development proceeds.

------------------------------------------------------------------------

### 5.7 Geographic Distribution of Clusters

```{r}
# Join cluster assignments to map data (use 2021 map, best quality)
map_clusters <- nuts2_map |>
  left_join(pc_scores |> select(geo, cluster), 
            by = c("NUTS_ID" = "geo"))

# Define cluster names
cluster_names_short <- c(
  "1" = "German Industrial (N=40, GDP €43k)",
  "2" = "Eastern Periphery (N=53, GDP €12.5k)",
  "3" = "Mediterranean Struggling (N=35, GDP €18k)",
  "4" = "Northern/Western Core (N=64, GDP €49k)",
  "5" = "Middle Class (N=62, GDP €29k)"
)

# Create map with colors matching 5.5 visualization
ggplot(map_clusters) +
  geom_sf(aes(fill = factor(cluster)), color = "white", size = 0.1) +
  scale_fill_manual(
    values = c(
      "1" = "#D73027",  # Red - German industrial
      "2" = "#4575B4",  # Blue - Eastern periphery
      "3" = "#ABDDA4",  # Green - Mediterranean struggling
      "4" = "#800080",  # Purple - Northern/Western core
      "5" = "#F46D43"   # Orange - Middle class
    ),
    name = "Regional Type",
    labels = cluster_names_short,
    na.value = "grey90"
  ) +
  coord_sf(xlim = c(-10, 35), ylim = c(35, 71)) +
  labs(
    title = "European Regional Typology: Five Cluster Solution",
    subtitle = "K-means clustering of PC1-PC3 scores (2021 data)",
    caption = "Source: Eurostat | Grey = No data or NUTS classification mismatch | Between-cluster variance: 72.3% | Portugal Map Data mismatching"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9, face = "bold"),
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    plot.caption = element_text(size = 7, hjust = 1)
  ) +
  guides(fill = guide_legend(nrow = 3, byrow = TRUE))
```

**Note on Portuguese Regions:** Three Portuguese mainland regions (Centro, Área Metropolitana de Lisboa, Alentejo) appear grey on the map due to NUTS classification changes between the data collection period and the reference map (NUTS 2021). The underlying data for these regions exists and was included in the PCA and clustering analysis; the grey shading reflects only cartographic incompatibility, not missing data.

**Cluster 4 (Purple): Northern & Western European Capitals and Regions** represents 25% of European regions (64 total), appearing as entire national territories such all of Scandinavia, the BeNeLux, Austria, part of Ireland, plus wealthy capitals (Warsaw, Paris, Prague, Madrid, Vilnius, Budapest) and regions within Spain (Basque Country) and France (Rhône-Alpes, Provence-Alpes, Midi Pirinées). This geographic concentration demonstrates that European prosperity extends beyond metropolitan cores, reflecting decades of regional developmental policies in social democratic welfare states.

**Cluster 1 (Red): The German Industrial Powerhouse** is geographically almost all of Germany. No other country's regions appear in this cluster besides two regions in Italy (Lombardia and Emilia-Romagna), highlighting Germany's distinctive economic model in Europe. This geographic isolation in clustering space mirrors Germany's singular role in European political economy as the continent's industrial backbone, maintaining manufacturing competitiveness through high productivity rather than transitioning to services as other wealthy nations have.

**Cluster 5 (Orange): European Middle Class** is the most geographically dispersed, appearing throughout the continent wherever moderate development has been achieved. Some eastern countries such as Estonia, some parts of Poland and the Sofia region in Bulgaria are present, otherwise it is scattered between Spanish/French/Italian regions. This dispersion reflects multiple pathways to middle-income status: post-socialist capitals integrating into EU markets, Spanish regions benefiting from EU structural funds, Italian regions with specialized manufacturing niches.

**Cluster 3 (Green): Mediterranean Struggling Regions** concentrates in Southern Europe—Greek regions (most of the country except Athens metropolitan area), Southern Spanish provinces, and Southern Italy (Mezzogiorno). The Mediterranean concentration reflects common challenges: tourism-dependent seasonality, debt crisis impacts (especially Greece), historical underdevelopment relative to Northern Europe, youth out-migration.

**Cluster 2 (Blue): Eastern European Periphery** is the old Eastern block. It is dominated by Bulgaria (all regions except Sofia), Romania (most regions except Bucharest), almost all Poland, Hungary, the Balkans region. This geographic concentration in post-socialist peripheries reflects incomplete transition—these regions experienced deindustrialization of inefficient socialist-era industry without attracting foreign investment or developing competitive services.

**Geographic Patterns:**

European development shows bimodal distribution: a large wealthy core (Cluster 4, 64 regions), substantial middle class (Cluster 5, 62 regions), and clear peripheries (Clusters 2-3, 88 regions). Within-country inequality often exceeds between-country gaps—Spain and Italy fragment across all five clusters, with Madrid resembling Stockholm more than rural Andalusia. Scandinavia's homogeneity (all regions Cluster 4) contrasts sharply with Southern/Eastern fragmentation, demonstrating that regional inequality reflects political choices rather than geographic inevitability. This bimodality creates political tension: enough wealthy regions could form degrowth coalitions, but substantial peripheries would resist.

### 5.8 Validation: ANOVA

```{r}
# Perform ANOVA for each variable
validation_data <- data |>
  select(-geo, -country) |>
  mutate(cluster = factor(pc_scores$cluster))

anova_results <- validation_data |>
  pivot_longer(cols = -cluster, names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarise(F_statistic = summary(aov(value ~ cluster))[[1]][["F value"]][1],
            p_value = summary(aov(value ~ cluster))[[1]][["Pr(>F)"]][1],
            .groups = "drop") |>
  arrange(desc(F_statistic)) |>
  mutate(significance = ifelse(p_value < 0.001, "***", 
                        ifelse(p_value < 0.01, "**", 
                        ifelse(p_value < 0.05, "*", "ns"))))

kable(anova_results, digits = c(0, 1, 4, 0),
      caption = "ANOVA: Do Clusters Differ Significantly?",
      col.names = c("Variable", "F-statistic", "p-value", "Sig."))
```

NOVA confirms clusters differ significantly (p \< 0.001) on all 15 variables (besides pop_density), validating the five-cluster solution identifies genuinely distinct regional types rather than arbitrary divisions.

**Life expectancy** (F=153.0) shows strongest differentiation—wealthy Cluster 4 lives 8-10 years longer than poor Cluster 2, a mortality gap comparable to differences between high-income and low-income countries globally, raising questions about EU cohesion.

**GHG emissions** (F=150.8) reveals a paradox: poorest regions (Cluster 2: 5,942 tons/capita) emit 40% less than German industrial regions, reflecting industrial concentration rather than sustainable practices—Eastern periphery's low emissions will rise with development unless green pathways are constructed.

**GDP per capita** (F=129.4) confirms the development gradient with nearly 4× difference between Cluster 4 (€45k) and Cluster 2 (€12.5k), exceeding gaps between some high- and middle-income countries internationally.

**Unemployment** (F=93.0) ranges from 3.6% (Cluster 1) to 16.0% (Cluster 3), indicating labor market conditions vary by an order of magnitude.

Even weaker differentiators remain significant: *renewables share* (F=18.3), *early school leaving* (F=13.6), *population density* (F=3.2, p=0.014). F-statistics ranging from 3.2 to 153.0 indicate clusters differ dramatically on core socioeconomic dimensions (life expectancy, GDP, emissions) while showing subtler but real differences on secondary characteristics, validating the typology as capturing meaningful regional variation.

### 5.9 Summary and Policy Implications

**Key Findings:**

K-determination methods conflicted (Silhouette: K=3, Gap: K=6, Elbow: K=5-6), but hierarchical clustering revealed five natural groupings. I selected K=5 as a pragmatic compromise. The solution explains 72.3% of between-cluster variance—exceeding PCA's 60%—suggesting discrete clustering captures additional structure beyond continuous gradients. Clusters range from 35-64 regions with no outliers, showing bimodal development: large wealthy core (Cluster 4: 64 regions, 25%), substantial middle (Cluster 5: 62 regions, 24%), combined peripheries (Clusters 2+3: 88 regions, 35%), and Germany's unique industrial model (Cluster 1: 40 regions, 16%). ANOVA confirms all clusters differ significantly on all 15 variables, with F-statistics ranging from 3.2 to 153.0.

**Differentiated Policy Implications:**

**Cluster 4 (Northern/Western Core, N=64):** Can pursue degrowth. GDP €49k, service-oriented, good environment (PC3=+0.94). Policies: consumption reduction (luxury taxes, work-time reduction), public goods investment, circular economy mandates, aviation restrictions.

**Cluster 1 (German Industrial, N=40):** Requires green industrial transformation, not degrowth. GDP €45k, but highest emissions. Policies: circular economy retrofitting, industrial energy efficiency, worker retraining, low-carbon procurement. Maintain manufacturing capacity while radically reducing emissions.

**Cluster 5 (Middle Class, N=62):** Needs selective growth—qualitative development without GDP expansion. GDP €29k. Policies: expand renewables, transit, education, healthcare while constraining automobiles, air travel, meat, fast fashion. Reallocate from harmful to beneficial sectors.

**Cluster 3 (Mediterranean Struggling, N=35):** Requires economic revitalization via sustainable pathways. GDP €18k, unemployment 16%, poverty 33%. Policies: just transition programs (care economy, renewable installation, move away from tourism-dependency), massive EU structural funds conditional on green development.

**Cluster 2 (Eastern Periphery, N=53):** Needs conventional development via green pathways. GDP €12.5k (lowest), poverty 23.7%. Degrowth impossible in existing scarcity. But possible with industrial transformations. Policies: infrastructure investment, educational expansion, agricultural modernization, green industrial development. Emissions will rise with development—ensure renewable energy, efficient buildings, sustainable agriculture.

**Political Economy Challenges:**

The nearly 4× GDP gap (Cluster 4: €49k vs. Cluster 2: €12.5k) and 8-10 year life expectancy difference represent fundamentally different realities. Degrowth advocacy from Stockholm/Vienna risks being perceived in rural Bulgaria/Romania as "eco-imperialism": wealthy regions impeding steps that themselves have used to achieve prosperity, and now are kept away from these poorer regions. Countering this requires: (1) massive EU redistribution ensuring Clusters 2-3 achieve material sufficiency before growth constraints apply, (2) differentiated timelines (Cluster 4 degrows now while redistributing wealth, Clusters 2-3 develop sustainably until reaching €30k+ thresholds), (3) just transition guarantees for carbon-dependent workers/communities.

Without differentiation, uniform policies will generate backlash, undermining EU legitimacy and environmental policy.

**Implications for Degrowth Theory:**

The "post-scarcity" precondition exists along a gradient—only Cluster 4 (25% of regions) that clearly meets it. Even Cluster 1 (wealthy industrial) and Cluster 5 (moderate) face structural barriers. This suggests degrowth may be appropriate for at most 25% of European regions, with alternative frameworks (green industrial policy, selective growth, sustainable development) more suitable for the remaining 75%. The finding that between-cluster variance (72.3%) exceeds continuous PC variance (60%) supports treating regional development as categorical types for policy purposes, though the overlapping ellipses we saw in section 5.5 remind us that boundaries remain fuzzy.

## 6. Conclusion

### 6.1 Summary of Findings

This analysis addressed: **What structural profiles characterize European NUTS-2 regions in terms of their "readiness" for a post-growth transition, and can we identify meaningful typologies across the continent?**

Principal Component Analysis of 15 socioeconomic, structural, and ecological variables across 254 regions identified three dominant dimensions explaining 60% of variance: **socioeconomic development** (PC1, 34%), **economic structure** (PC2, 15%), and **environmental performance** (PC3, 11%). K-means clustering revealed **five distinct regional types** explaining 72.3% of between-cluster variance—notably exceeding the 60% captured by continuous PC scores. ANOVA validation confirmed all clusters differ significantly (p\<0.001) on all 15 variables, with F-statistics ranging from 3.2 to 153.0.

The five clusters range from the Northern/Western European Core (Cluster 4, N=64, GDP €49k) to the Eastern European Periphery (Cluster 2, N=53, GDP €12.5k) a 4× difference comparable to gaps between high- and lower-middle-income countries globally. Geographic analysis revealed bimodal rather than normal distribution: large wealthy core (25%), substantial middle (24%), and clear peripheries (35%), with Germany and Northern Italy's unique industrial model (16%) standing apart. Within-country inequality often exceeds between-country gaps—Spain and Italy fragment across four to five clusters.

**Critically**, "readiness" as operationalized here captures **material preconditions**—economic slack, institutional capacity—not political will or current practices. The analysis reveals *who could* pursue degrowth without welfare collapse, not *who wants to* or *currently does*. The "readiest" regions (Cluster 4) are simultaneously the highest consumers, underscoring that material capacity and ecological performance are closely linked dimensions.

### 6.2 Policy Implications

Uniform EU-wide policies would be inappropriate. Each cluster requires differentiated approaches:

**Cluster 4:** Can pursue degrowth experimentation. Policies: consumption reduction (luxury taxes, work-time reduction), public goods investment, circular economy mandates, aviation restrictions.

**Cluster 1:** Requires green industrial transformation, not degrowth. Policies: circular economy retrofitting, industrial energy efficiency, worker retraining, low-carbon procurement. Maintain manufacturing while radically reducing emissions.

**Cluster 5:** Needs selective growth—qualitative development without GDP expansion. Policies: expand renewables, transit, education, healthcare while constraining automobiles, air travel, meat, fast fashion.

**Cluster 3:** Requires economic revitalization via sustainable pathways. Policies: just transition programs (care economy, renewable installation, sustainable tourism), massive EU structural funds conditional on green development.

**Cluster 2:** Needs conventional development via green pathways. Policies: infrastructure investment, educational expansion, agricultural modernization, green industrial development. Current low emissions will rise—ensure renewable energy, efficient buildings, sustainable agriculture.

**The Political Economy Challenge:** The 4× GDP gap and 8-10 year life expectancy difference represent fundamentally different realities. Degrowth advocacy could be perceived in rural areas as "eco-imperialism" from the EU and its powerful members.

### 6.3 Limitations

**Operationalization of "Readiness":** This analysis operationalizes "degrowth readiness" as material capacity, not political will or current practices.

**Environmental Variables' Geographic Aggregation:** PC3 relies on three country-level variables unavailable at NUTS-2 level. This means PC3 captures **national environmental policies** rather than regional variation. When stating "Cluster 4 has good environmental performance," this means "Cluster 4 regions are disproportionately in countries with strong national performance," not that these specific regions have low emissions. Future research with regional emissions inventories would enable more granular analysis.

**Cross-Sectional Limitations:** The 2021 snapshot misses dynamics. Some low-capacity regions may be rapidly developing; others trapped in persistent underdevelopment. Additionally, 2021's post-COVID context potentially affects employment composition, though most variables are slow-moving.

**PC1 Replicates HDI:** The first component strongly correlates with existing development indices. While theoretically justified via post-scarcity logic, empirically PC1 is "HDI renamed." The contribution lies in showing how development gradients interact with industrial structure and environmental performance to create distinct types requiring differentiated policies.

### 6.4 Future Research

**Longitudinal Analysis:** Track how regions move through typological space. Are Cluster 5 regions converging toward Cluster 4, or diverging? Panel data (2010-2025) would reveal convergence/divergence dynamics.

**Regional Environmental Data:** As regional emissions inventories become available, re-run analysis with true regional-level ecological variables to test whether PC3's link holds.

**Political Attitudes:** Complement structural analysis with survey data on public attitudes toward growth. Do Cluster 4 populations actually support degrowth policies? Testing whether material capacity predicts normative orientation would validate the "post-scarcity" hypothesis.

**Policy Experiments:** Evaluate regions attempting degrowth-adjacent policies (Vienna's housing, Amsterdam / London's doughnut economics, Barcelona's superblocks). What enables success or causes failure?

### 6.5 Final Reflections

European regional development is fundamentally heterogeneous, characterized by discrete subregions with its own characteristics rather than smooth gradients. The core finding: **only 25% of regions (Cluster 4) meet the "post-scarcity" precondition** degrowth theorists identify as necessary. Even wealthy industrial (Cluster 1) and moderate (Cluster 5) regions face mejor structural and political barriers. This suggests degrowth as currently theorized may be appropriate for at most one-quarter of European regions, with alternative frameworks more suitable for the remainder.

The political economy challenge is quite ambicious: wealthy regions advocating degrowth must confront "eco-imperialism" critiques from peripheric regions experiencing their demands as development constraints that wealthy regions themselves did use. Any viable European degrowth politics requires massive redistribution, honest demands, and a just transition guaranteed.

Perhaps the analysis's deepest insight is this: **regional inequality within Europe is stronger than international inequality**. The €49k vs. €12.5k GDP gap and 8-10 year life expectancy gap mean "Europe" is not a coherent unit for policy purposes. These are numbers that differentiate a so-called "first world country" to a "third world country. With these premises is impossible to even think of implementing a single policy-for-all approach in Europe. Madrid resembles Stockholm more than it resembles rural Andalusia; Bavaria differs from rural Bulgaria as much as Milan does to Sicily.

The question is not whether Europe should pursue degrowth, but **which Europes** should pursue which pathways toward ecological sustainability, human wellbeing, and a different kind of economics that work in the interest of the people rather then in the interests of economics for its sake. The five-cluster typology provides a framework for answering this question, translating abstract degrowth theory into concrete, differentiated policy interventions appropriate to diverse material realities.
